Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-15 csf-2017/paper.tex
added a couple of words to prevent line break in reference causing hyperlink to break
Some work on discussion and RW.

\textbf{Formulating control flow integrity} The attentive reader may notice
that, while we claim that our stack and return pointer discipline enforces
control flow integrity, we do not prove a general theorem to show that this is
the case. The reason for this is that it is not clear what such a theorem should
look like. Formulations in terms of a control flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.\footnote{It is a
  bit unfortunate that the general term ``control flow integrity'' is used for a
  property that only captures partial correctness of control flow.} In fact, our
examples show that our logical relation imply a form of control flow integrity
that is stronger than such formulations, although this is not made very
explicit. As future work, we are considering to look at a more explicit and
useful way to formalise well-bracketed control flow. The idea would be to define
a variant of our capability machine with built-in call and return instructions
and well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our stack and return pointer
discipline is fully abstract~\cite{abadi_protection_1998}, which roughly means
that the lower-level machine does not allow interactions with the compiled code
that do not correspond to an interaction with the original code in the
high-level machine.

\textbf{Performance and the requirement for stack clearing} It is important to
consider the performance of our proposed stack and return management discipline.
Generally, the additional security measures described in
Section~\ref{sec:stack-and-return-pointer} obviously impose an overhead w.r.t.
an approach that does not provide security. However, they generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements that we have for stack clearing, in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear the entire stack,
except the part that we are using ourselves. This means that on every crossing
of the boundary between trusted and adversary code, a potentially large region
of memory needs to be zeroed out entirely. We believe that capability machines
like CHERI should consider to provide special support for this requirement: a
highly-optimized instruction for zeroing out a potentially large block of
memory. We have no experience in processor design, but we imagine this might be
optimized as a vector instruction or even further, by having the processor keep
track of a range of memory for which loads should pretend that it has already
been cleared, by returning zero irrespective of the actual contents of the
memory.

\textbf{Modularity} Our return pointers and callbacks satisfy the checks we
apply to the ones we get from the adversary (return pointers are allowed to be
local, callbacks must be global), i.e.\ we do not claim special status and what
we propose is modular.

\textbf{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to features that have been proposed for
high-level languages. Specifically, \citet{osvald_gentrification_2016}'s
second-class or local values in Scala are remarkably similar: they are a kind of
values that can be provided to other code for immediate use without allowing it
to store the value in a closure or reference for later use. We believe reasoning
about such values will require very similar reasoning techniques than the ones
we have used for local capabilities.

@@

Our logical relation is a unary, step-indexed Kripke logical relation (SKLR)
with recursive
worlds~\cite{Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by \citet{Devriese:2016ObjCap} to formulate
capability safety in a high-level JavaScript-like lambda calculus. Our
fundamental theorem is similar to theirs and can be seen as a formulation of
capability safety of the capability machine. Because we are not interested in
externally observable side-effects (like console output or memory access
traces), we do not require their notion of effect parametricity.

The idea of applying SKLRs to an assembly language was applied before by
\citet{Hur:2011:KLR:1926385.1926402}, who constructed a binary SKLR for
formulating and proving correctness of an ML-to-assembly compiler. Their
condition on the assembly code are quite loose (for example, they allow
hand-written and self-modifying assembly code), but they do assume code that is
non-adversarial and respects the calling convention of the high-level language
(for example, the code does not store return pointers and invoke them after
their intended lifetime). As a result, they can treat return pointers as
evaluation contexts (continuations) in a lambda calculus and use biorthogonality
to define the LR~\cite{pitts_operational_1998}. In our setting, there is no
guarantee that potentially adversarial code will respect the calling convention,
so, as described in Section~\ref{sec:logical-relation}, we do not use
biorthogonal closure but just model return pointers as continuations using a
single orthogonal closure.

Another novel aspect of our logical relation is how we model the temporary
nature of local capabilities. The main insight is that the temporary, revokable
nature of these capabilities can be seen as a generalisation of the
syntactically special, unstoreable status of evaluation contexts in lambda
calculi (without control effects). Closely related to this is a study by
\citet{Dreyer:jfp12} (DNB) of the impact of higher-order state and control
effects on local relational reasoning. DNB formally capture the special status
of evaluation contexts using Kripke worlds with public and private future world
relations. Our future world relations are inspired by theirs and they are at
times remarkably similar. For example, the state-transitition system in the
proof of our assembly version of the awkward example, is exactly the same as the
one used by DNB for the original awkward example. However, in our logical
relation, public and private future worlds turn up in an entirely different
place in the logical relation: rather than using public future worlds for the
special syntactic category of evaluation contexts, they are used in the value
relation depending on the depending on the locality of the capability at hand.
Additionally, our worlds are a bit more complex because, to allow store-local
capabilities, they can have temporary regions that are only monotonous w.r.t.
public future worlds, while DNB's worlds are entirely permanent.

%/comparison with related Dreyer-Neis-Birkedal and with Hur-Dreyer}
%\todo[inline]{Add discussion/comparison to other LRs}
% Comparisson of LR to other work:
% - Single othorgonal closure
%   o all capabilities are possible continuations, not just the one we
%   intend to be it.
% - No termination guarantees in observation relation
% - Private/public transitions used like in Dreyer-Neis-Birkedal, but
%   also to  handle the difference in what local and global
%   capabilities can handle.

% As already mentioned, our the state transition systems used in our
% logical relation are inspired by earlier work on logical relations for
% high-level
% languages~\cite{Ahmed:popl09, Dreyer:jfp12,Devriese:2016ObjCap}.
% The idea of public and private transitions comes from 
% Dreyer et. al.~\cite{Dreyer:jfp12}, who gave a logical relations model
% of a high-level lambda calculus. In \emph{loc. cit.}
% public transitions were used to 
% capture the \emph{extensional} behavior of a program expression
% (and a context), whereas private transitions were used to capture
% the internal behaviour of a computation using local state.
% Technically, the logical relation of Dreyer et. al.\ was
% formulated using biorthogonal closure, and the future world
% relation for continuations was required to be a public future world
% relation. In other words, the choice between private or public future
% rested on the ``structure'' of the programming language. 
% In contrast, 
% for our low-level, unstructured, assembly language, we
% we instead decide on whether to use a private or public future 
% world depending on whether the capability is global or local. 
% Another interesting technical difference is that in our low-level 
% language we do not have a fixed calling convention and any register
% can serve as a continuation, and therefore we do not use biorthogonal
% closure but just model return pointers as continuations using a single
% orthogonal closure.

% \lars{don't know yet whether we need something like the following paragraph:}
% Even though 
% Hur and Dreyer~\cite{Hur:2011:KLR:1926385.1926402} also
% considered a logical
% relation involving an assembly language, they still used
% biorthogonal closure but that was because their logical relation
% was a relation between a high-level language and an assembly
% language,  and thus the assembly language in the relation essentially
% inherited structure from the high-level language.

>>>>

\subsection{Stack and return pointer discipline}
\textbf{Formulating control flow integrity} The attentive reader may notice
that, while we claim that our stack and return pointer discipline enforces
control flow integrity, we do not prove a general theorem to show that this is
the case. The reason for this is that it is not clear what such a theorem should
look like. Formulations in terms of a control flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.\footnote{It is a
  bit unfortunate that the general term ``control flow integrity'' is used for a
  property that only captures partial correctness of control flow.} In fact, our
examples show that our logical relation imply a form of control flow integrity
that is stronger than such formulations, although this is not made very
explicit. As future work, we are considering to look at a more explicit and
useful way to formalise well-bracketed control flow. The idea would be to define
a variant of our capability machine with built-in call and return instructions
and well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our stack and return pointer
discipline is fully abstract~\cite{abadi_protection_1998}, which roughly means
that the lower-level machine does not allow interactions with the compiled code
that do not correspond to an interaction with the original code in the
high-level machine.

\textbf{Performance and the requirement for stack clearing} It is important to
consider the performance of our proposed stack and return management discipline.
Generally, the additional security measures described in
Section~\ref{sec:stack-and-return-pointer} obviously impose an overhead w.r.t.
an approach that does not provide security. However, they generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements that we have for stack clearing, in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear the entire stack,
except the part that we are using ourselves. This means that on every crossing
of the boundary between trusted and adversary code, a potentially large region
of memory needs to be zeroed out entirely. We believe that this in fact is a
common requirement for typical usage scenario's of local capabilities and
capability machines like CHERI should consider to provide special support for
this requirement, in the form of a highly-optimized instruction for zeroing out
a potentially large block of memory. We have no experience in processor design,
but we imagine this might be optimized as a vector instruction or even further,
by having the processor keep track of a range of memory for which loads should
pretend that it has already been cleared, by returning zero irrespective of the
actual contents of the memory.

\textbf{Modularity} An important aspect of our stack and return pointer
discipline is that it is modular. By that, we mean that the requirements we have
on callbacks, return pointers etc. provided to trusted code by an adversary are
also satisfied by callbacks, return pointers etc. that we construct ourselves.
For example, our return pointers are necessarily local capabilities because they
must point to memory where we can store the old stack pointer, but the
adversary's return pointers are also allowed to be local. Adversary callbacks on
the other hand are required to be global but the callbacks we construct
ourselves are also allocated on the heap and global. In other words, we do not
assume that our code has some sort of special status and it is possible for the
adversary to protect themselves against us in the same way as we protect against
them.

\textbf{Single orthogonal closure} The definitions of $\stder$ and $\stdvr$ in
Figure~\ref{fig:logrel} apply a single orthogonal closure, a new variant of an
existing pattern called biorthogonality. Biorthogonality is a pattern for
defining logical relations~\citep{krivine_classical_1994,pitts_operational_1998}
in terms of an observation relation of safe configurations (like we do). The
idea is to define safe evaluation contexts as the set of contexts that produce
safe observations when plugging safe values and define safe terms as the set of
terms that can be plugged into safe evaluation contexts to produce safe values.
This is an alternative to more direct definitions where safe terms are defined
as terms that evaluate to safe values. An advantage of biorthogonality is that
it scales better to languages with control effects like call/cc. Our definitions
can be seen as a variant of biorthogonality, where we take only a single
orthogonal closure: we do not define safe evaluation contexts but immediately
define safe terms as those that produce safe observations when plugged with safe
values. This is natural because we model arbitrary assembly code that does not
necessarily respect a particular calling convention. As a result, return
pointers are in principle values like all others and there is no reason to treat
them specially in the logical relation.

It is interesting to note that \citet{Hur:2011:KLR:1926385.1926402} also build a
step-indexed, Kripke, biorthogonal logical relation for an assembly language
(for reasoning about correct compilation from ML to assembly), but because they
only model non-adversarial code that treats return pointers specially according
to a particular calling convention, they can use standard biorthogonality rather
than a single orthogonal closure like us.

\textbf{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary nature of local capabilities using public/private
future worlds. The main insight is that the temporary, revokable nature of these
capabilities can be seen as a generalisation of the syntactically-enforced
unstoreable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\citet{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalising the idea that it is safe to make temporary internal state
modifications (private world transititions, that invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning (i.e. invoking the continuation). We
generalise this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to \citet{Dreyer:jfp12}'s (for example, our proof of
the awkward example uses exactly the same state transition system than theirs),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and store-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\textbf{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \citet{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing it to be stored in a closure or reference for
later use. We believe reasoning about such values will require very similar
reasoning techniques than the ones we have used for local capabilities.

@@

Our logical relation is a unary, step-indexed Kripke logical relation (SKLR)
with recursive
worlds~\cite{Appel:2001:IMR:504709.504712,Ahmed2004semantics,Birkedal:2011:SKM:1926385.1926401},
closely related to the one used by \citet{Devriese:2016ObjCap} to formulate
capability safety in a high-level JavaScript-like lambda calculus. Our
fundamental theorem is similar to theirs and can be seen as a formulation of
capability safety of the capability machine. Because we are not interested in
externally observable side-effects (like console output or memory access
traces), we do not require their notion of effect parametricity. Our logical
relation uses several ideas from previous work, like Kripke worlds with regions
containing state transition systems~\citep{Ahmed:popl09}, public/private future
worlds~\citep{Dreyer:jfp12} (see Section~\ref{sec:discussion} for a discussion),
biorthogonality~\citep{pitts_operational_1998,benton_biorthogonality_2009-1,Hur:2011:KLR:1926385.1926402}.
